---
title: "M cell organoid digital PCR analysis"
author: "Rachel Creighton"
date: "2026-07-15"
output:
  html_document:
    keep_md: true
---
```{r setup, message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(here)
```

# Overview
In this project, we used digital PCR to measure the expression of M cell marker genes (GP2, SpiB, CCL20), interferon-stimulated genes (ISG15, MX1, IFI6, etc), and TLRs. 

We initially iterated through several previously published organoid media formulations to determine which resulted in the highest M cell marker expression. These experiments were performed in duodenal-derived organoids from a single donor.

# M cell marker expression
## Base media comparison
The first set of experiments, evaluating "differentiation" media published by Ding et al and "patterning" and "maturation" media published by He et al were repeated 3 times.

```{r get data, message=FALSE}
experiments <- paste("organoid5", "organoid6", "organoid7", sep = "|")
file_names <- list.files(path = here("data", "dpcr", "raw"),
                         pattern = experiments)

data <- list()
for (i in 1:length(file_names)) {
  tmp <- read_csv(here("data", "dpcr", "raw", file_names[i])) %>%
    mutate(experiment_name = str_extract(file_names[i], "^\\d{4}_\\d{4}_\\w{9}")) %>%
    rename(chamber = "ChamberID",
          sample_desc = "SampleName",
          droplet_num = "TotalNumberOfDroplets",
          blue_conc = "Blue_Channel_Concentration",
          green_conc = "Green_Channel_Concentration",
          experiment_name = "experiment_name") %>%
    mutate(chamber = str_sub(chamber, -2)) %>%
    select(chamber, sample_desc, droplet_num, blue_conc, green_conc, experiment_name) %>%
    pivot_longer(c(blue_conc, green_conc), names_to = "color", names_pattern = "(.*)_conc", values_to = "concentration")
  
  data[[i]] <- tmp
}

tidy_data <- bind_rows(data)
```

```{r get metadata, message=FALSE}
experiments <- paste("organoid5", "organoid6", "organoid7", sep = "|")
file_names <- list.files(path = here("data", "dpcr", "metadata"),
                         pattern = experiments)

metadata <- list()
for (i in 1:length(file_names)) {
  tmp <- readxl::read_xlsx(here("data", "dpcr", "metadata", file_names[i])) %>%
    mutate(experiment_name = str_extract(file_names[i], "^\\d{4}_\\d{4}_\\w{9}")) %>%
    mutate(sample = as.character(sample)) %>%
    pivot_longer(c(blue, green), names_to = "color", values_to = "gene")
  
  metadata[[i]] <- tmp
}

tidy_metadata <- bind_rows(metadata)
```

```{r combine data and metadata}
tidy_full_data <- full_join(tidy_data, tidy_metadata) %>%
  filter(droplet_num >= 8000) %>%
  filter_out(sample == "NTC" | sample == "NRT") %>%
  drop_na() %>%
  mutate(orig_concentration = concentration * dilution_factor) %>%
  mutate(organoid_experiment = str_extract(experiment_name, "organoid\\d+$"))

head(tidy_full_data)
```
```{r normalize data}
normalized_data <- tidy_full_data %>%
  dplyr::select(c(organoid_experiment, sample, orig_concentration, gene)) %>%
  filter_out(orig_concentration == "Inf") %>%
  pivot_wider(names_from = gene, values_from = orig_concentration, values_fn = mean) %>%
  mutate(across(.cols = c("CCL20", "GP2", "SPIB"),
                .fns = ~ .x / RPP30,
                .names = "{.col}_norm"))

```

