---
title: "M cell organoid digital PCR analysis"
author: "Rachel Creighton"
date: "2026-07-15"
output:
  html_document:
    keep_md: true
---
```{r setup, message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(here)
```

# Overview
In this project, we used digital PCR to measure the expression of M cell marker genes (GP2, SpiB, CCL20), interferon-stimulated genes (ISG15, MX1, IFI6, etc), and TLRs. 

We initially iterated through several previously published organoid media formulations to determine which resulted in the highest M cell marker expression. These experiments were performed in duodenal-derived organoids from a single donor.

# M cell marker expression
## Base media comparison
The first set of experiments, evaluating "differentiation" media published by Ding et al and "patterning" and "maturation" media published by He et al were repeated 3 times.


```{r get data}
experiments <- paste("organoid5", "organoid6", "organoid7", sep = "|")
file_names <- list.files(path = here("data", "dpcr", "raw"),
                         pattern = experiments)
file_names
```

